---
title: "Haiti Comparisons 2.0"
author: "Danny"
date: "11/06/2020"
output:
  html_document:
    fig_width: 12
    fig_height: 6
  pdf_document: default
---
## summary

```{r packages, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(here)
library(ggplot2)
library(lubridate)
library(reshape2)
library(viridis)
library(RColorBrewer)
library(tidyr)
library(hydroGOF)
library(stringr)
library(knitr)
library(kableExtra)
library(rnaturalearthdata)
library(rnaturalearth)
library(ggrepel)
library(sp)
library(tibble)
library(verification)
library(purrr)
library(dplyr)
library (ggspatial)
library(sf)
```

```{r setup, include=FALSE}
source(here("src", "helper_funs.R"))
zm <- read.csv(here( "data", "alpha_grouped.csv"))
zm <- dplyr::filter(zm, station != "thiotte")
zm$date <- as.Date(zm$date)
view(zm)

zm <- transmute(zm,
                station = station,
                latitde = latitude,
                longitude = longitude, 
                date = date, 
                year = year(date),
                month = month(date),
                day = day(date),
                rain = rain,
                chirps_rain = chirps_rain_adj
)

view(zm)
if(anyDuplicated(zm %>% dplyr::select(station, date))) stop("Duplicates found!")


# 1 Jan = 1
s_doy_start <- 1
zm <- zm %>% mutate(doy = yday_366(date),
                    s_doy = (doy - s_doy_start + 1) %% 366,
                    s_doy = ifelse(s_doy == 0, 366, s_doy),
                    syear = ifelse(s_doy > (366 - s_doy_start + 1), year - 1, year),
                    month_abb = month(date, label = TRUE))
```

```{r monthly_zeros}
# Look at months with zero total rainfall and no missing values

zm_zero_months <- zm %>%
  group_by(station, month,year) %>%
  summarise(t_rain = sum(rain, na.rm = TRUE),
            n_na = sum(is.na(rain))) %>%
  filter(t_rain == 0 & n_na == 0)%>%

write.csv(here ("data", "zero_months_adj.csv"))
zm_month <- zm %>%
  group_by(station, year, month) %>%
  summarise(t_rain = sum(rain),
            n_na = sum(is.na(rain)))

zm_month_wide <- zm_month %>%
  pivot_wider(id_cols = c(year, month), names_from = station, values_from = t_rain)

zm <- zm %>%
  # Group by station and month for the calculation
  group_by(station, year, month) %>%
  # Add two columns for total monthly rainfall and number of missing values in the month
  mutate(t_rain = sum(rain, na.rm = TRUE), 
         # for each station and each month we calculate the total rain
         n_na = sum(is.na(rain)),
         # for each station and each month we calculate the number of missing values
         # Make rain_adj = NA if monthly total was 0 and there were no missing values
         # and otherwise rain_adj = rain
         # Some of these could be true zeros and could be checked later
         rain_adj = ifelse(t_rain == 0 & n_na == 0, NA, rain))

zm$rain <- zm$rain_adj
zm$rain_adj <- NULL
zm$t_rain <- NULL
zm$n_na <- NULL

zm <- zm %>%
  mutate(high_rain = rain > 200,
         rain = ifelse(high_rain, NA, rain),
         chirps_rain = ifelse(high_rain, NA, chirps_rain))
zm$high_rain <- NULL
```

```{r data_formats}
zm_long_st <- zm %>% 
  melt(id.vars = c("station", "date", "year", "syear", "month", "month_abb", 
                   "doy", "s_doy"),
       measure.vars = names(zm)[endsWith(names(zm), "rain")][-1],
       variable.name = "product", value.name = "pr_rain")

#the melt function takes data in a wide range of formats and stacks a set of columns into a single column of data

#creates a rainday column where the values are true or false depending on if there is rain that day or not
zm_long <- zm %>% 
  melt(id.vars = c("station", "date", "year", "syear", "month", "month_abb", "doy", "s_doy"),
       measure.vars = names(zm)[endsWith(names(zm), "rain")],
       variable.name = "product", value.name = "rain") %>%
  mutate(rainday = rain > 1)

zm_long$product <- recode(zm_long$product, rain = "station")
stations <- c( "cap_haitien", "quanaminthe", "jeremie", "damien", "petitionville", 
               "jacmel", "les_cayes") #north to south would be a better order for these 
products <- levels(zm_long$product)
products <- products[-1]
names(products) <- substr(products, 1, nchar(products) - 5)

metadata_station <- read.csv(here("data", "daily_metadata.csv"))

metadata_station$station <- factor(metadata_station$station, levels = stations) #making the 

zm_long$station <- factor(zm_long$station, levels = stations)

by_station <- zm_long %>%
  group_by(station) %>%
  filter(!is.na(rain)) %>%
  summarise(first_date = first(date), #this line of code seems to be giving the same last date for all our data 
            last_date = last(date))

metadata_station <- left_join(metadata_station, by_station, by = "station")
rm(by_station)

skable <- function(kable_input) {
  kable_input %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                                full_width = FALSE)
}
```

## Stations

```{r}
metadata_station %>% 
  dplyr::select(station, latitude, longitude) %>%   ##may have to add a min year/max year column
  kable(digits = 2) %>%
  skable()
```


```{r station_map, fig.width=9, fig.height=4.5}
hti_admbnda_adm0_cnigs_20181129 <- 
  sf::st_read(quiet=TRUE, 
              dsn = here("data", "hti_adm_cnigs_20181129",  
                         "hti_admbnda_adm0_cnigs_20181129.shp")) %>%
  dplyr::mutate(lon = purrr::map_dbl(geometry, ~sf::st_centroid(.x)[[1]]),
                lat = purrr::map_dbl(geometry, ~sf::st_centroid(.x)[[2]]))

locs <- read.csv(here("data", "daily_metadata.csv"))
width <- 0.05

ggplot2::ggplot() +
  ggplot2::geom_sf(data = hti_admbnda_adm0_cnigs_20181129, 
                   mapping=ggplot2::aes()) + 
  theme_grey() +
  geom_point(data = locs, aes(x = longitude, y = latitude), colour = "black") +
  
 geom_text_repel(data = locs, aes(x = var1, y = var2, label = station, fontface = "italic"), 
                 colour = "black", size = 3) 
```

```{r gof_fun}
dgof <- function(df, sim, obs, na.rm = TRUE) {
  g <- hydroGOF::gof(sim = df[[sim]], obs = df[[obs]], na.rm = na.rm) # how this function is working? 
  x <- as.list(g)
  names(x) <- row.names(g)
  x
}

comp_stats <- c("r", "ME", "PBIAS %", "MAE", "NSE", "rSD")
names(comp_stats) <- c("Correlation coefficient (1 = Perfect)",
                       "Mean bias (same units)",
                       "Percentage bias (%)",
                       "Mean absolute bias (same units)",
                       "Nash-Sutcliffe efficiency (1 = Perfect)",
                       "Ratio of standard deviations (< 1 less variable, > 1 more variable)")

comp_stats_digits <- c(2, 0, 0, 0, 2, 3)
```

## Yearly Comparisons 

```{r yearly_calcs}
quality_control <- zm_long %>% 
  group_by(syear,station,month) %>% 
  summarise(
    monthly_rain = sum(rain)
    
  )

by_syear <- zm_long %>%
  group_by(syear,station,product) %>%
  summarise(total_rain = sum(naif_nmin(rain, 330)),
            n_rain = sum(naif_nmin(rain, 330) > 0.1),
            max_rain = max(naif_nmin(rain, 330)),
            mean_rain = total_rain/n_rain,
            n_na = sum(is.na(rain))
)

by_syear_st <- by_syear %>% 
  pivot_wider(id_cols = c(station, syear),
              names_from = product, values_from = total_rain:n_na, names_sep = "__") 

gof_syear <- by_syear_st %>%  
  group_by(station) %>%
  nest() %>%
  mutate(n = purrr::map_int(data, 
                       ~sum(!is.na(.$total_rain__station) &
                              !is.na(.$total_rain__chirps_rain))), 
         #calculates number of complete years we have 
         gof__total_rain = purrr::map(data, dgof, "total_rain__chirps_rain", 
                                      "total_rain__station", 
                                      na.rm = TRUE),
         gof__n_rain = purrr::map(data, dgof, "n_rain__chirps_rain", "n_rain__station", 
                                  na.rm = TRUE),
         gof__mean_rain = purrr::map(data, dgof, "mean_rain__chirps_rain", 
                                     "mean_rain__station", 
                                     na.rm = TRUE) 
        
  )

gof_pr <- gof_syear %>% 
  unnest(cols = data)
```

## hurricane season comparisons 


```{r hurricane_season_calc}
zm_long_h <- filter(zm_long,month %in% c(6:11))

by_syear_h <- zm_long_h %>%
  group_by(syear,station,product) %>%
  summarise(total_rain = sum(naif_nmin(rain, 150)),    #since we are dealing with 6 month here instead of the full year i had to change from 330 to 150 
            n_rain = sum(naif_nmin(rain, 150) > 0.1),
            max_rain = max(naif_nmin(rain, 150)),
            mean_rain = total_rain/n_rain,
            n_na = sum(is.na(rain))
)

by_syear_st_h <- by_syear_h %>% 
  pivot_wider(id_cols = c(station, syear),
              names_from = product, values_from = total_rain:n_na, names_sep = "__") 

gof_syear_h <- by_syear_st_h %>%  
  group_by(station)%>%
  nest() %>%
  mutate(n = purrr::map_int(data, 
                       ~sum(!is.na(.$total_rain__station) &
                              !is.na(.$total_rain__chirps_rain))), 
         gof__total_rain = purrr::map(data, dgof, "total_rain__chirps_rain", 
                                      "total_rain__station", 
                                      na.rm = TRUE),
         gof__n_rain = purrr::map(data, dgof, "n_rain__chirps_rain", "n_rain__station", 
                                  na.rm = TRUE),
         gof__mean_rain = purrr::map(data, dgof, "mean_rain__chirps_rain", 
                                     "mean_rain__station", 
                                     na.rm = TRUE) 
        
  )

gof_pr_h <- gof_syear_h %>% 
  unnest(cols = data)


```


```{r yearly_plots_fun}
yearly_plots <- function(df, gof_col, stat_pr, stat_st) {
  max_y <- max(c(df[[stat_pr]], df[[stat_st]]), na.rm = TRUE)
  dat <- df %>% 
    pivot_longer(cols = c(stat_pr, stat_st), names_to = "product", values_to = stat_pr) %>%
    mutate(ME = purrr::map_dbl(.data[[gof_col]], "ME"),
           r = purrr::map_dbl(.data[[gof_col]], "r"),
           rSD = purrr::map_dbl(.data[[gof_col]], "rSD")
           )
  mean_df <- dat %>% 
    group_by(station, product) %>% 
    summarise(m = mean(.data[[stat_pr]], na.rm = TRUE))
  g <- ggplot(dat, aes(x = syear, y = .data[[stat_pr]], colour = product)) +
    geom_line() +
    geom_point() +
    geom_hline(data = mean_df, aes(yintercept = m, colour = product)) +
    scale_x_continuous(limits = c(1979, 2012)) +
    # n
    geom_text(data = dat, aes(label = paste("n", n)), size = 4,
              x = 1979, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    # bias
    geom_text(data = dat, aes(label = paste("bias", signif(ME, 2))), 
              size = 4, x = 1979 + 6, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    # cor
    geom_text(aes(label = paste("cor", round(r, 2))), 
              size = 4, x = 1979 + 14, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    # rSD
    geom_text(aes(label = paste("rSD", round(rSD, 2))), 
              size = 4, x = 1979 + 22, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    ggtitle(paste(stat_pr, ":", "CHIRPS", "vs", "station")) +
    facet_wrap(~station)
  g
}
```

```{r stats_tables_fun}
stats_tables <- function(df, obj_col, obj_stats = comp_stats, name) {
  dat <- df %>% 
    ungroup()
  for (i in seq_along(obj_stats)) {
    dat <- dat %>%
      mutate(purrr::map_dbl(df[[obj_col]], obj_stats[i]))
    names(dat)[ncol(dat)] <- obj_stats[i]
  }
  dat <- dat %>% 
    mutate(station = as.character(station))
  dat <- dat[ , c("station", obj_stats)]
  dat %>%
  kable(digits = comp_stats_digits[i], caption = paste("Comparison statistics:", name), 
        format.args = list(big.mark = ",")) %>%
  skable() %>%
  print()
}
```

### Comparison Statistics for Total Yearly Rainfall

```{r yearly_total_plots}
yearly_plots(gof_pr, gof_col = "gof__total_rain", 
                        stat_pr = "total_rain__chirps_rain", 
                        stat_st = "total_rain__station")

ggsave(here("results", "syear_total_rain_adj.png"), width = 12, height = 6)

```


```{r yearly_n_obs, results="asis"}
gof_syear %>% 
  dplyr::select(station, n) %>%
  kable(caption = "Number of years compared") %>%
  skable()
```

```{r yearly_total_rain_tables, results="asis"}
stats_tables(gof_syear, "gof__total_rain", name = "Total Yearly Rainfall")
```

### Comparison Statistics for Number of rainy days

```{r yearly_n_rain_plots}

yearly_plots(gof_pr, gof_col = "gof__n_rain", 
             
                        stat_pr = "n_rain__chirps_rain", 
                        stat_st = "n_rain__station")
ggsave(here("results", "syear_n_rain_adj.png"), width = 12, height = 6)
#want to change the y axis to read n_rain instead of chirps
```

```{r yearly_n_rain_tables, results="asis"}
stats_tables(gof_syear, "gof__n_rain", name = "number of rain days ")
```

## hurricane season comparisons (total rainfall)
``` {r hurricane_season_plots}

yearly_plots(gof_pr_h, gof_col = "gof__total_rain", 
                        stat_pr = "total_rain__chirps_rain", 
                        stat_st = "total_rain__station")

ggsave(here("results", "syear_total_rain_h_adj.png"), width = 12, height = 6)

```

```{r yearly_total_rain_tables_h, results="asis"}
stats_tables(gof_syear_h, "gof__total_rain", name = "Total Yearly Rainfall")
```

``` {r hurricane_season_n_rain_plots }
yearly_plots(gof_pr_h, gof_col = "gof__n_rain", 
             
                        stat_pr = "n_rain__chirps_rain", 
                        stat_st = "n_rain__station")

ggsave(here("results", "syear_n_rain_h_adj.png"), width = 12, height = 6)

```

```{r yearly_n_rain_tables_h, results="asis"}
stats_tables(gof_syear_h, "gof__n_rain", name = "number of rain days ")
```

## Monthly Comparisons

```{r monthly_calcs}
by_month <- zm_long %>%
  group_by(station, syear, month_abb, product) %>%
  #filter(month %in% c(11:12, 1:4)) %>%
  summarise(total_rain = sum(naif_nmin(rain, 25)),  
            n_rain = sum(naif_nmin(rain, 25) > 0.1),  
            max_rain = max(naif_nmin(rain, 20)), 
            mean_rain = ifelse(n_rain == 0, NA, total_rain/n_rain), 
            n_na = sum(is.na(rain))
            )

by_month_st <- by_month %>%
  pivot_wider(id_cols = c(station, syear, month_abb), 
              names_from = product, values_from = total_rain:mean_rain, names_sep = "__") 


gof_month <- by_month_st %>%
  group_by(station) %>%
  nest() %>%
  mutate(n = purrr::map_int(data, 
                        ~sum(!is.na(.$total_rain__station) & 
                               !is.na(.$total_rain__chirps_rain))),
         gof__total_rain = purrr::map(data, dgof, "total_rain__chirps_rain", 
                                      "total_rain__station", 
                                      na.rm = TRUE),
         gof__n_rain = purrr::map(data, dgof, "n_rain__chirps_rain", "n_rain__station", 
                                  na.rm = TRUE),
         gof__mean_rain = purrr::map(data, dgof, "mean_rain__chirps_rain", 
                                     "mean_rain__station", 
                                      na.rm = TRUE)
         
         )


gof_pr_month <- gof_month %>%
  unnest(cols = data)
```

```{r monthly_plots_fun}
monthly_plots <- function(df, stat_pr,gof_col, stat_st, product_name) {
  max_y <- max(c(df[[stat_pr]], df[[stat_st]]), na.rm = TRUE)
  vals_relace <- c(product_name, "station")
  names(vals_relace) <- c(stat_pr, stat_st)
  dat <- df %>% 
    filter(month_abb %in% month.abb[c(1:12)]) %>%
    pivot_longer(cols = c(stat_pr, stat_st), names_to = "product", values_to = stat_pr) %>%
    mutate(product = recode(product, !!!vals_relace),
           ME = purrr::map_dbl(.data[[gof_col]], "ME"),
           r = purrr::map_dbl(.data[[gof_col]], "r"),
           rSD = purrr::map_dbl(.data[[gof_col]], "rSD"))
  mean_df <- dat %>% 
    group_by(station, product, month_abb) %>% 
    summarise(m = mean(.data[[stat_pr]], na.rm = TRUE))

  g <- ggplot(dat, aes(x = syear, y = .data[[stat_pr]], colour = product)) +
    geom_line() +
    geom_point() +
    geom_hline(data = mean_df, aes(yintercept = m, colour = product)) +
    scale_x_continuous(limits = c(1979, 2012)) +
    ggtitle(paste(stat_pr, ":", product_name, "vs", "station")) +
    facet_wrap(~month_abb) 
     # n
  #  geom_text(data = dat, aes(label = paste("n", n, sep = "  ")), size = 4,
              #x = 1979, y = max_y, na.rm = TRUE, 
             # inherit.aes = FALSE)  +
    
    # bias
   # geom_text(data = dat, aes(label = paste("bias", signif(ME, 2), sep = "  ")), 
            #  size = 4, x = 1979 + 6, y = max_y, na.rm = TRUE, 
           #   inherit.aes = FALSE) +
    # cor
    #geom_text(aes(label = paste("cor", round(r, 2), sep = "  ")), 
          #    size = 4, x = 1979 + 14, y = max_y, na.rm = TRUE, 
         #     inherit.aes = FALSE)  +
    # rSD
    #geom_text(aes(label = paste("rSD", round(rSD, 2), sep = "  ")), 
     #         size = 4, x = 1979 + 22, y = max_y, na.rm = TRUE, 
        #      inherit.aes = FALSE) 
    
    
  g
}

```

### Comparison statistics for total monthly rainfall

```{r monthly_plots_total_rain}
for (i in seq_along(stations)) {
  s <- stations[i]
  monthly_plots(gof_pr_month %>% filter(station == s),
                stat_pr = "total_rain__chirps_rain", gof_col = "gof__total_rain",
                stat_st = "total_rain__station", product_name = "CHIRPS")
  
  ggsave(here( "results", paste0("month_total_rain_adj", s, ".png")), width = 12, height = 6)
}
```

```{r monthly_tables_total_rain, results="asis"}
stats_tables(gof_month, "gof__total_rain", name = "total_rain_stats")
```

### Comparison statistics for number of monthly raindays


```{r monthly_plots_n_rain}
for (i in seq_along(stations)) {
  s <- stations[i]
monthly_plots(gof_pr_month %>% filter(station == s), gof_col = "gof__n_rain",
                        stat_pr = "n_rain__chirps_rain", 
                        stat_st = "n_rain__station", product_name = "CHIRPS")  

ggsave(here("results", paste0("smonth_n_rain_adj", s, ".png")), width = 12, height = 6)
}

```

```{r monthly_tables_n_rain, results="asis"}
stats_tables(gof_month, "gof__n_rain", name = "n_rain_stats_month")
```
## Markov Chain models of the chance of rain


```{r markov_chain_setup}
haiti_markov <- zm %>% 
  filter(!is.na(chirps_rain) & !is.na(rain)) %>%
  mutate(rainday1 = rain > 0.85,
         pr_rainday1 = chirps_rain > 0.85,
         pr_rainday2 = chirps_rain > 2,
         pr_rainday3 = chirps_rain > 3,
         pr_rainday4 = chirps_rain > 4,
         pr_rainday5 = chirps_rain > 5)

f_zero_order_station <- rainday1 ~ (cos(s_doy * 1 * 2 * pi/366) +
                                    sin(s_doy * 1 * 2 * pi/366) + 
                                    cos(s_doy * 2 * 2 * pi/366) + 
                                    sin(s_doy * 2 * 2 * pi/366) +
                                    cos(s_doy * 3 * 2 * pi/366) +
                                    sin(s_doy * 3 * 2 * pi/366))
f_zero_order_product <- update.formula(f_zero_order_station, pr_rainday1 ~ .)

predict_stack_lst <- list()

for(s in seq_along(stations)) {  
  predict_df <- data.frame(station = stations[s], s_doy = 1:366,
                           s_doy_date = as.Date(1:366, origin = as.Date("1999/12/31")))
  dat <- haiti_markov %>%
    filter(station == stations[s])

  zero_order_station <- glm(f_zero_order_station, data = dat, family = binomial)
  zero_order_product <- glm(f_zero_order_product, data = dat, family = binomial)
  #print(anova(zero_order_station, test="Chisq"))
  predict_df[["station_fit"]] <- predict(zero_order_station, newdata = predict_df,
                                         type = "response")
  predict_df[["chirps_fit"]] <- predict(zero_order_product, newdata = predict_df,
                                        type = "response")
  
  f_zero_order_product_2thres <- update.formula(f_zero_order_station, pr_rainday2 ~ .)
  f_zero_order_product_3thres <- update.formula(f_zero_order_station, pr_rainday3 ~ .)
  f_zero_order_product_4thres <- update.formula(f_zero_order_station, pr_rainday4 ~ .)
  f_zero_order_product_5thres <- update.formula(f_zero_order_station, pr_rainday5 ~ .)
  fms_thres <- list(f_zero_order_product_2thres, f_zero_order_product_3thres,
                    f_zero_order_product_4thres, f_zero_order_product_5thres)
  for(j in seq_along(fms_thres)) {
    zero_order <- glm(fms_thres[[j]], data = dat, family = binomial)
    predict_df[[paste0("chirps", "_", j + 1, "thres")]] <- predict(zero_order,
                                                                   newdata = predict_df,
                                                                   type = "response")
  }

  predict_stack <- predict_df %>% melt(id.vars = c("station", "s_doy", "s_doy_date"), 
                                       variable.name = "product", value.name = "prob")

  predict_stack$product <- as.character(predict_stack$product)
  predict_stack_lst[[length(predict_stack_lst) + 1]] <- predict_stack
  # Plot small amounts
  # g <- ggplot(predict_stack, aes(x = s_doy, y = prob, colour = type)) +
  #   geom_line() +
  #   facet_wrap(~product2) +
  #   scale_color_manual(values = c("black", c25[1:7])) +
  #   ggtitle(paste("Chance of rain:", stations[s]))
  # ggsave(here("results", "haiti", paste0("haiti_", "markov_zero", stations[s], ".png")), 
  #        plot = g, width = 12, height = 6)
}
predict_stack_all <- bind_rows(predict_stack_lst)
predict_stack_all$product <- factor(predict_stack_all$product)
predict_stack_all$product <- relevel(predict_stack_all$product, "station_fit")
```


```{r markov_chain_plots}
  dat <- predict_stack_all %>% filter(product %in% c("station_fit", "chirps_fit"))
  g <- ggplot(dat, aes(x = s_doy_date, y = prob, colour = product, size = product)) +
    geom_line() +
    facet_wrap(~station) +
    scale_size_manual(values = c(0.8, rep(0.6, 4))) +
    scale_color_manual(values = c("black", c25[1:4])) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b") +
    ggtitle(paste("Chance of rain"))
  print(g)
  ggsave(here("results", paste0("markov_zero_adj.png")),
         plot = g, width = 12, height = 6)
```

```{r markov_chain_plots_thresh}
  dat <- predict_stack_all
  g <- ggplot(dat, aes(x = s_doy_date, y = prob, colour = product, size = product)) +
    geom_line() +
    facet_wrap(~station) +
    scale_size_manual(values = c(0.8, rep(0.6, 5))) +
    scale_color_manual(values = c("black", c25[1:5])) +
    scale_x_date(date_breaks = "2 months", date_labels = "%b") +
    ggtitle(paste("Chance of rain"))
  print(g)
  ggsave(here( "results", paste0("markov_zero_threshs_adj.png")),
         plot = g, width = 12, height = 6)
```



